---
title: "Project EDA"
author: "Savannah Sauers, Lera Martinez, Simon Tao"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
```

```{r, include=FALSE}
project <- read.csv("~/Desktop/Fall 2026/DATS6101/DATS6101_Project_MTA_Subway_Hourly_Ridership.csv")
```

```{r}
# Make sure columns are cast to correct types
project$borough <- as.factor(project$borough)
project$payment_method <- as.factor(project$payment_method)
project$fare_class_category <- as.factor(project$fare_class_category)
project$ridership <- as.numeric(project$ridership)
project$transfers <- as.numeric(project$transfers)
```

```{r}
head(project)
str(project)
```

## Clean Data (create columns that will be used for our analysis)
```{r}
# Add 24-hour time column
project <- project %>%
  mutate(
    timestamp = mdy_hms(transit_timestamp),
    hour = hour(timestamp)
  )
```

```{r}
# Add categories for morning, afternoon, evening, night
# Morning (5am - noon), afternoon (noon - 4pm), evening (4pm - 8pm), night (8pm - 5am)
project <- project %>%
  mutate(
    time_of_day = case_when(
      hour >= 5 & hour < 12 ~ "Morning",
      hour >= 12 & hour < 16 ~ "Afternoon",
      hour >= 16 & hour < 20 ~ "Evening",
      TRUE ~ "Night"
    )
  )
```


```{r}
# Add day of week column
project <- project %>%
  mutate(
    day_of_week = wday(timestamp, label = TRUE)
  )
```

```{r}
head(project)
```

## EDA
```{r}
length(unique(project$station_complex_id))
```

There are 424 stations in the dataset.

```{r}
# Week average for total ridership by borough
borough_ridership_means <- project %>%
  group_by(borough, transit_timestamp) %>%
  summarize(hourly_ridership = sum(ridership, na.rm=TRUE), .groups = "drop") %>%
  group_by(borough) %>%
  summarize(mean_week_ridership = mean(hourly_ridership, na.rm=TRUE))

borough_ridership_means
```

Greatest to least mean total ridership by borough: Manhattan, Brooklyn, Queens, Bronx

This shows the average number of riders in each borough. Includes all stations and times of day.

```{r}
# Week average hourly ridership by station and borough
station_ridership_means <- project %>%
  group_by(borough, station_complex_id) %>%
  summarize(mean_ridership = mean(ridership, na.rm=TRUE)) %>%
  arrange(desc(mean_ridership))

station_ridership_means
```

Out of the top 10 stations for mean hourly ridership, 7 of them are in Manhattan, 2 of them are in Queens, and 1 of them is in Brooklyn

```{r}
scatter <- project %>%
  ggplot(aes(x=ridership, y=transfers, color=borough)) +
  geom_smooth(method = "lm")

scatter
```

## Hypothesis Testing
Two-sample t-test
- We do not know our population standard deviation
- Heavily right skewed, but sample size is large enough

# ANOVA: Differences in ridership between time-of-day categories

# ANOVA: Differences in hourly ridership between boroughs
```{r}
# Daily ridership total for each borough
borough_daily <- project %>%
  mutate(date = as.Date(timestamp)) %>%
  group_by(date, borough) %>%
  summarize(total_ridership = sum(ridership, na.rm = TRUE), .groups = "drop")
```

```{r}
# ANOVA test: Do daily averages in each borough differ?
borough_anova <- aov(total_ridership ~ borough, data = borough_daily)
summary(borough_anova)
```

```{r}
borough_post_hoc <- TukeyHSD(borough_anova)
borough_post_hoc
```
