---
title: "Project EDA"
author: "Savannah Sauers, Lera Martinez, Simon Tao"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r, include=FALSE}
library(tidyverse)
```

```{r, include=FALSE}
project <- read.csv("~/Desktop/Fall 2026/DATS6101/DATS6101_Project_MTA_Subway_Hourly_Ridership.csv")ß
```

```{r}
# Make sure columns are cast to correct types
project$borough <- as.factor(project$borough)
project$payment_method <- as.factor(project$payment_method)
project$fare_class_category <- as.factor(project$fare_class_category)
project$ridership <- as.numeric(gsub(",", "", project$ridership))
project$transfers <- as.numeric(gsub(",", "", project$transfers))
```

# Proofing data
```{r Checking-Data}
#Data exported corresponds to the week defined in the project
range(project$timestamp, na.rm = TRUE)

#Checking for any NA values
colSums(is.na(project))

#Checking for duplicates 
sum(duplicated(project))

```

```{r 24h-Graph-prelim}
#Using the mean, proposal/skeleton of potential graph that 
#maps over the entire week the average ridership over a 24h period
hourly_ridership <- project %>%
  group_by(hour) %>%
  summarize(
    total_ridership = sum(ridership, na.rm = TRUE),
    mean_ridership = mean(ridership, na.rm = TRUE),
    sd_ridership = sd(ridership, na.rm = TRUE)
  )

day_mean_graph <- ggplot(hourly_ridership, aes(x = hour, y = mean_ridership)) +
  geom_line(color = "#1A5279",linewidth = 1) +
  geom_point(color = "#2980B9",size = 1) +
  labs(
    title = "Average NYC Subway Ridership by Hour",
    x = "Hour of Day",
    y = "Average Ridership"
  ) +
  theme_minimal()+ 
  theme(panel.grid.minor = element_blank())

day_mean_graph
```

```{r}
head(project)
str(project)
```

## Clean Data (create columns that will be used for our analysis)
```{r}
# Add 24-hour time column
project <- project %>%
  mutate(
    timestamp = mdy_hms(transit_timestamp),
    hour = hour(timestamp)
  )
```

```{r}
# Add categories for morning, afternoon, evening, night
# Morning (5am - noon), afternoon (noon - 4pm), evening (4pm - 8pm), night (8pm - 5am)
project <- project %>%
  mutate(
    time_of_day = case_when(
      hour >= 5 & hour < 11 ~ "Morning",
      hour >= 11 & hour < 16 ~ "Mid Day",
      hour >= 16 & hour < 21 ~ "Evening",
      TRUE ~ "Night"
    )
  )
```

```{r}
# Add day of week column
project <- project %>%
  mutate(
    day_of_week = wday(timestamp, label = TRUE)
  )
```

```{r}
# Weekend vs. Weekday column
project <- project %>%
  mutate(day_type = if_else(day_of_week %in% c("Sat", "Sun"), "Weekend", "Weekday"))
```

```{r}
str(project)
```

## EDA
```{r}
length(unique(project$station_complex_id))
```

There are 424 stations in the dataset.

```{r}
# Week average for total ridership by borough
borough_ridership_means <- project %>%
  group_by(borough, transit_timestamp) %>%
  summarize(hourly_ridership = sum(ridership, na.rm=TRUE), .groups = "drop") %>%
  group_by(borough) %>%
  summarize(mean_week_ridership = mean(hourly_ridership, na.rm=TRUE))

borough_ridership_means
```

Greatest to least mean total ridership by borough: Manhattan, Brooklyn, Queens, Bronx

This shows the average number of riders in each borough. Includes all stations and times of day.

```{r}
# Week average hourly ridership by station and borough
station_ridership_means <- project %>%
  group_by(borough, station_complex_id) %>%
  summarize(mean_ridership = mean(ridership, na.rm=TRUE)) %>%
  arrange(desc(mean_ridership))

station_ridership_means
```

Out of the top 10 stations for mean hourly ridership, 7 of them are in Manhattan, 2 of them are in Queens, and 1 of them is in Brooklyn

```{r}
scatter <- project %>%
  ggplot(aes(x=ridership, y=transfers, color=borough)) +
  geom_smooth(method = "lm")

scatter
```

## Hypothesis Testing
Questions of interest:

- “How do NYC subway ridership volumes vary throughout the day during the week of September 15-21, 2024?” -- Anova and post hoc

- Which hours have the highest ridership average (weekdays vs. weekends)? 

- Do these hours correspond to commuting periods? 
  
**T-Tests: Commute vs. Non-commute; Weekday vs. Weekend**
```{r}
hourly_ridership <- project %>%
  mutate(commute_period = if_else(time_of_day %in% c("Morning", "Evening"), "Commute","Non-Commute")) %>%
  group_by(timestamp, day_type, time_of_day, commute_period) %>%
  summarize(total_ridership = sum(ridership, na.rm = TRUE), .groups = "drop")
```

Commute vs. Non-commute time periods:
```{r}
qqnorm(hourly_ridership$total_ridership[hourly_ridership$commute_period == "Commute"])
qqline(hourly_ridership$total_ridership[hourly_ridership$commute_period == "Commute"])
```

```{r}
qqnorm(hourly_ridership$total_ridership[hourly_ridership$commute_period == "Non-Commute"])
qqline(hourly_ridership$total_ridership[hourly_ridership$commute_period == "Non-Commute"])
```

Weekday vs. Weekend periods:
```{r}
qqnorm(hourly_ridership$total_ridership[hourly_ridership$day_type == "Weekday"])
qqline(hourly_ridership$total_ridership[hourly_ridership$day_type == "Weekday"])
```

```{r}
qqnorm(hourly_ridership$total_ridership[hourly_ridership$day_type == "Weekend"])
qqline(hourly_ridership$total_ridership[hourly_ridership$day_type == "Weekend"])
```

Assumptions check for t-tests:

- Commute vs. Non-commute time
  - We do not know the population SD (we took a sample of a typical week with no holidays or off season periods)
  - Sample size is large enough (per the CLT) for the test to be robust against normality violations
  - Independent observation assumption may be violated due to time data, but continue with t-test for this class

- Weekday vs. Weekend
  - We do not know the population SD (we took a sample of a typical week with no holidays or off season periods)
  - Sample size is large enough (per the CLT) for the test to be robust against normality violations
  - Independent observation assumption may be violated due to time data, but continue with t-test for this class
  
```{r}
# t-test for commute time vs. non-commute time means
commute_period_test <- t.test(total_ridership ~ commute_period, data = hourly_ridership)
commute_period_test
```

Null hypothesis: Mean hourly ridership of commute periods = Mean hourly ridership of non-commute periods

Alternative hypothesis: Mean hourly ridership of commute periods ≠ Mean hourly ridership of non-commute periods

Alpha = 0.05 > P-value = 2.061e-10 --> Reject the null hypothesis. There is sufficient evidence to conclude that the mean hourly ridership during commute periods is different that the mean hourly ridership during non-commute

```{r}
# t-test for weekday vs. weekend means
weekday_weekend_test <- t.test(total_ridership ~ day_type, data = hourly_ridership)
weekday_weekend_test
```

Null hypothesis: Mean hourly ridership of Weekdays = Mean hourly ridership of weekends

Alternative hypothesis: Mean hourly ridership of Weekdays ≠ Mean hourly ridership of weekends

Alpha = 0.05 > P-value = 7.411e-06 --> Reject the null hypothesis. There is sufficient evidence to conclude that the mean hourly ridership on weekdays is different that the mean hourly ridership of weekends

**ANOVA: Differences in ridership between time-of-day categories**

ANOVA Assumptions Check:
- Normality:
- Homogeneity of variances:
- Independence of errors:

```{r}
time_anova <- aov(total_ridership ~ time_of_day, data = hourly_ridership)
summary(time_anova)
```

Null hypothesis: Mean hourly ridership during morning = Mean hourly ridership during midday = Mean hourly ridership during evening = Mean hourly ridership during night

Alternative hypothesis: The mean hourly ridership of at least one time of day category is different than the others.

Alpha = 0.05 > P-value = <2e-16 --> Reject the null hypothesis. There is sufficient evidence to conclude that at least one time category group has a different mean hourly ridership than the others.

```{r}
time_post_hoc <- TukeyHSD(time_anova)
time_post_hoc
```

Post hoc conclusions:
- At the alpha = 0.05 level, the time periods pairings that have a significant difference are:
  - Midday and Evening
  - Morning and Evening
  - Night and evening
  - Night and midday
  - Night and morning
- Out of these, I think the notable differences in pairs are midday and evening and morning and evening (it makes sense that night is significantly lower than the others).

**ANOVA: Differences in hourly ridership between boroughs**

ANOVA Assumptions Check:
- Normality:
- Homogeneity of variances:
- Independence of errors:

```{r}
# Daily ridership total for each borough
borough_daily <- project %>%
  mutate(date = as.Date(timestamp)) %>%
  group_by(date, borough) %>%
  summarize(total_ridership = sum(ridership, na.rm = TRUE), .groups = "drop")
```

```{r}
# ANOVA test: Do daily averages in each borough differ?
borough_anova <- aov(total_ridership ~ borough, data = borough_daily)
summary(borough_anova)
```

Null hypothesis: Mean daily ridership in Manhattan = Mean daily ridership in Brooklyn = Mean daily ridership in Queens = Mean daily ridership in the Bronx

Alternative hypothesis: The mean daily ridership of at least one borough is different than the others.

Alpha = 0.05 > P-value = <2e-16 --> Reject the null hypothesis. There is sufficient evidence to conclude that at least one borough has a different mean daily ridership than the other boroughs.

```{r}
borough_post_hoc <- TukeyHSD(borough_anova)
borough_post_hoc
```

Post hoc conclusions:
- At the alpha = 0.05 level, the borough pairings that have a significant difference are:
  - Brooklyn and the Bronx
  - Manhattan and the Bronx
  - Manhattan and Brooklyn
  - Queens and Manhattan
- Manhattan's mean daily ridership differs significantly from the mean daily ridership in all other boroughs

# Transfers EDA, total and average bus to station transfers
```{r, message=FALSE}
project %>%
  group_by(station_complex, station_complex_id) %>%
  summarize(
    total_transfers = sum(transfers, na.rm = TRUE),
    average_transfers = mean(transfers, na.rm = TRUE),
    observations = n()
  ) %>%
  arrange(desc(total_transfers))
```

```{r}
top_transfer_stations <- project %>%
  group_by(station_complex_id, borough) %>%
  summarize(
    total_transfers = sum(transfers, na.rm = TRUE),
    .groups = "drop"
  ) %>%
  arrange(desc(total_transfers)) %>%
  slice_head(n = 10)

ggplot(top_transfer_stations, aes(x = reorder(station_complex_id, total_transfers), y = total_transfers, fill = borough)) + geom_col()
```

The top five total transfer stations are Flushing-Main St (7), Jamaica Center-Parsons/Archer (E,J,Z), Kew Gardens-Union Tpke (E,F), 74-Broadway (7)/Jackson Hts-Roosevelt Av (E,F,M,R), Jamaica-179 St (F). There is a drop in total transfers after the Jamaica-179 St.

The five highest transfer stations for the week of Sept 15-21 are also located in Queens.

#Distribution of transfers at top 10 stations by hour
```{r}
hourly_transfers <- project %>%
  semi_join(
    top_transfer_stations,
    by = c("station_complex_id")
  ) %>%
  group_by(station_complex_id, hour) %>%
  summarize(
    mean_transfers = mean(transfers, na.rm = TRUE),
    .groups = "drop"
  )
ggplot(hourly_transfers, aes(x = hour, y = mean_transfers, color = factor(station_complex_id), group = station_complex_id)) + geom_line() + labs(
  title = "Average Hourly Transfers at Top 10 Transfer Stations",
  x = "Hour of Day",
  y = "Mean Number of Transfers",
  color = "Station Number"
)
```
All 10 stations show a spike in average transfers during the morning commute time. Flushing-Main St. Station shows an elevated transfer rate through the day and evening.
